Modelling biological pathway dynamics\\with \tas

### 2013-07-01 Paper.tex
Minor textual adjustments

Systems biology aims at understanding biological systems from a holistic rather than a reductionist perspective.
As the complexity of dynamic molecular interaction networks exceeds the capacity of the human mind, 
computational support is needed to describe and understand such networks.
Large volumes of data are particularly useful when used as reference for functional models
that formalize the interactions between the components of a network.
Such models can assist in exploration and transfer of knowledge. 
Functional disease models can be combined with patient-specific expression profiles to stratify 
patients into treatment groups and maximize efficacy.

@@

The reaction rate $r$ is computed from a single kinetic constant $k$ and the activity levels of the reactants. Three kinetic scenarios are implemented :\\
%\begin{description}
{\bf Scenario 1}: $r(a) = k \times a$ : the reaction rate depends on the activity level of the upstream enzyme only. \\
{\bf Scenario 2}: $r(a, b) = k \times a \times b$ : the reaction rate depends on the activity levels of both the upstream enzyme and the
downstream substrate. $b$ represents the inactive fraction of substrate if the reaction
is an activation, whereas it refers to the active fraction when the effect is inhibitory.\\
{\bf Scenario 3}: $r(c, d) = k \times c \times d$ : the reaction rate depends on the activity of two upstream components chosen by the user. This scenario performs a function analogous to an AND-gate in Boolean models.

@@

It has been observed that the activation of extracellular regulated kinase (ERK) shows significantly different 
dynamics upon treatment with EGF as compared to treatment with NGF. A transient, peak-shaped 
activation is observed when PC-12 cells are treated with EGF, whereas treatment with NGF results in sustained activation of ERK,
even after removal of the input signals via growth factor-neutralizing antibodies. These activation profiles are tied to different cellular 
outcomes, with transient activation leading to proliferation and sustained activation leading to differentiation. 
These findings led the authors of~\cite{egf-ngf} to hypothesize the existence of a positive feedback from ERK to an upstream node in the network and 
existence of a new network component that blocks this feedback. It was found that NGF prevents this blocking, leading to sustained activation~\cite{egf-ngf}.
We have formalized and implemented these topological reasonings in an ANIMO model, shown in Figure~\ref{fig:case-study-model}. 
The parameters used in this network are given in Table~\ref{tab:case-study-parameters}.
The reference experimental conditions are the addition of either EGF or NGF
in sufficient quantity to fully activate their respective receptors. After 10 minutes a growth factor-neutralizing 
antibody is added, shutting off the input signal to the network. The evolution of the network is
observed for 60 minutes from the initial treatment.

@@

Tools that allow to construct models based on ODEs include for example COPASI~\cite{copasi}, 
E-Cell~\cite{e-cell} and GNA~\cite{gna}.
These tools add to the potential of ODEs by coupling them with additional modelling approaches,
such as stochastic models\footnote{Stochastic modelling becomes particularly useful when some
molecular species have very small concentrations, which nullifies the assumption of a well-mixed solution used in models based on ODEs.}
used in COPASI and E-Cell, and by allowing for qualitative modelling, as does GNA.
Other tools, such as Systems Biology Workbench~\cite{sbw}, allow to couple high-performance
analysis engines (like roadRunner) to user-friendly interfaces (e.g. JDesigner). Finally, tools like CellDesigner~\cite{celldesigner}
provide an appealing graphical representation as an interface to powerful computational engines.
Most of the cited tools are founded on parameter-intensive approaches, while the paradigm implemented in ANIMO
uses less precise formulations, which requires less initial knowledge from the user.

@@

The use of \tas\ as the underlying formalism of ANIMO lies the foundation to the use of more advanced 
analysis techniques to query constructed models. In particular, applying model checking techniques would allow us to perform \emph{in silico} experiments, answering questions
such as ``What is the combination of inputs that leads to $\mbox{activity}(A) \geq 20/50$
and $\mbox{activity}(B) < 10/80$ in 120 minutes?''. Moreover, taking advantage of statistical model
checking capabilities recently introduced to UPPAAL~\cite{uppaal-smc}, we plan to extend the current modelling paradigm adding the possibility
to define stochastic behaviour (e.g., assign an exponential distribution to the addition of a drug as input to the network),
and support probabilistic queries such as ``What is the probability that
$\mbox{activity}(A) \geq 40/50$ in 30 minutes?''.
Other interesting developments are aimed at further speeding up the modelling phase, in order to let the user
start interrogating a model as soon as possible. We plan to include a support for parameter
sensitivity analysis and automated parameter fitting to a given experimental data set. 
After having defined a measure of distance between the model and the experimental data, we plan to minimize that distance
via (user-defined) parameter sweeps.
We are also developing
techniques based on automata learning~\cite{test-based-modelling} for deriving the topology of a biological network, based
on a series of constraints and experimental data series.

>>>>

Systems biology aims at understanding biological systems from a holistic rather than a reductionist perspective.
As the complexity of dynamic molecular interaction networks exceeds the capacity of the human mind, 
computational support is needed to describe and understand such networks.
Large volumes of data are particularly useful when used as reference for functional models
that formalize the interactions between the components of a network.
Such models can assist in exploration and transfer of knowledge. 
Functional disease models can be combined with patient-specific expression profiles to stratify 
patients into groups and maximize treatment efficacy.

@@

The reaction rate $r$ is computed from a single kinetic constant $k$ and the activity levels of the reactants ($a, b$). Three kinetic scenarios are implemented :\\
%\begin{description}
{\bf Scenario 1}: $r(a) = k \times a$ : the reaction rate depends on the activity level of the upstream enzyme only. \\
{\bf Scenario 2}: $r(a, b) = k \times a \times b$ : the reaction rate depends on the activity levels of both the upstream enzyme and the
downstream substrate. $b$ represents the inactive fraction of substrate if the reaction
is an activation, whereas it refers to the active fraction when the effect is inhibitory.\\
{\bf Scenario 3}: $r(c, d) = k \times c \times d$ : the reaction rate depends on the activity of two upstream components chosen by the user. This scenario performs a function analogous to an AND-gate in Boolean models.

@@

It has been observed that the activation of extracellular regulated kinase (ERK) shows significantly different 
dynamics upon treatment with EGF as compared to treatment with NGF. A transient, peak-shaped 
activation is observed when PC-12 cells are treated with EGF, whereas treatment with NGF results in sustained activation of ERK,
even after removal of the input signals via growth factor-neutralizing antibodies. These activation profiles are tied to different cellular 
outcomes, with transient activation leading to proliferation and sustained activation leading to differentiation. 
These findings led the authors of~\cite{egf-ngf} to hypothesize the existence of a positive feedback from ERK to an upstream node in the network and 
existence of a new network component that blocks this feedback. It was found that NGF prevents this blocking, leading to sustained activation~\cite{egf-ngf}.
We have formalized and implemented these topological reasonings in an ANIMO model, shown in Figure~\ref{fig:case-study-model}. 
The parameters used in this network are given in Table~\ref{tab:case-study-parameters}.
The accompanying experimental conditions are the addition of either EGF or NGF
in sufficient quantity to fully activate their respective receptors. After 10 minutes a growth factor-neutralizing 
antibody is added, shutting off the input signal to the network. The evolution of the network is
observed for 60 minutes from the initial treatment.

@@

Tools that allow to construct models based on ODEs include for example COPASI~\cite{copasi}, 
E-Cell~\cite{e-cell} and GNA~\cite{gna}.
These tools add to the potential of ODEs by coupling them with additional modelling approaches,
such as stochastic models\footnote{Stochastic modelling becomes particularly useful when some
molecular species have very small concentrations, which nullifies the assumption of a well-mixed solution used in models based on ODEs.}
used in COPASI and E-Cell, and by allowing for qualitative modelling, as does GNA.
Other tools, such as Systems Biology Workbench~\cite{sbw}, allow to couple high-performance
analysis engines (such as roadRunner) to user-friendly interfaces (e.g. JDesigner). Finally, tools such as CellDesigner~\cite{celldesigner}
provide an appealing graphical representation as an interface to powerful computational engines.
Most of the cited tools are founded on parameter-intensive approaches, while the paradigm implemented in ANIMO
uses less precise formulations, which requires less initial knowledge from the user.

@@

The use of \tas\ as the underlying formalism of ANIMO lies the foundation to the use of more advanced 
analysis techniques to query constructed models. In particular, applying model checking techniques would allow us to perform \emph{in silico} experiments, answering questions
such as ``What is the combination of inputs that leads to $\mbox{activity}(A) \geq 20/50$
and $\mbox{activity}(B) < 10/80$ in 120 minutes?''. Moreover, taking advantage of statistical model
checking capabilities recently introduced to UPPAAL~\cite{uppaal-smc}, we plan to extend the current modelling paradigm adding the possibility
to define stochastic behaviour. This could for instance be done by assigning exponential distributions to a set of reactions,
to support probabilistic queries such as ``What is the probability that
$\mbox{activity}(A) \geq 40/50$ in 30 minutes?''.
Other interesting developments are aimed at further speeding up the modelling phase, in order to let the user
start interrogating a model as soon as possible. We plan to include a support for parameter
sensitivity analysis and automated parameter fitting to a given experimental data set. 
After having defined a measure of distance between the model and the experimental data, we plan to minimize that distance
via user-defined parameter sweeps.
We are also developing
techniques based on automata learning~\cite{test-based-modelling} for deriving the topology of a biological network, based
on a series of constraints and experimental data series.
